---
title: "Data Manipulation and Analysis with R"
author: Michael J. Moon
date: September 8, 2019
output:
  ioslides_presentation:
    css: style.css
---

# R Basics

```{r include=FALSE}
library(knitr)
library(kableExtra)
library(RColorBrewer)
library(scales)
palette(brewer.pal(8, 'Accent'))
```

## What is R?

From `https://www.r-project.org/about.html`...  
    
<div class="quotation">
R is a language and environment for statistical computing and graphics...R provides a wide variety of statistical (...) and graphical techniques, and is highly extensible...R provides an <b>Open Source</b> route...R is available as <b>Free</b> Software under the terms of the Free Software Foundation’s GNU General Public License in source code form...
</div>


## What is R?

From `https://www.r-project.org/about.html`...  
    
<div class="quotation">
R is a language and environment for statistical computing and graphics...R provides a wide variety of statistical (...) and graphical techniques, and is highly extensible...R provides an <b>Open Source</b> route...R is available as <b>Free</b> Software under the terms of the Free Software Foundation’s GNU General Public License in source code form...
</div>

### Open Source and free

+ No money? No problem.
+ You can also contribute to the community.

## What is R?

From `https://www.r-project.org/about.html`...  
    
<div class="quotation">
R is a **language** and **environment** for statistical computing and graphics...R provides a wide variety of statistical (...) and graphical techniques, and is highly extensible...
</div>

### A language

+ Allows you to **instruct** computers to perform statistical tasks. 
+ Needs to learn the language.

### An environment

+ You need the `R` environment so your computer can understand the `R` instructions.
+ `R` installation comes with a basic source code editor.

## RStudio

### RStudio

+ A popular IDE dedicated to `R`.
+ Provides an integrated GUI for source code editing, plotting, and debugging among other features.
+ This document was created in `RStudio`.

### Installation

+ Download `R` from [https://www.r-project.org/](https://www.r-project.org/).
+ Download `RStudio` from [https://www.rstudio.com/products/rstudio/download/](https://www.rstudio.com/products/rstudio/download/)  
    (Use *RStudio Desktop* free version).
+ Install `R` before installing `RStudio`.

## Data types
### Basic types

<div class="columns-2">
+ Numeric (real numbers)
```{r eval=FALSE}
1/7 
```
+ Integer
```{r eval=FALSE}
30 
```
+ Complex
```{r eval=FALSE}
3+2i 
```
+ Logical
```{r eval=FALSE}
30==30 
# TRUE <- this is a comment
```
+ Character (not a number)
```{r eval=FALSE}
as.character(3.14) 
# "3.14" not a number
```
</br>
</div>

## Data types

### Not allowed

```{r eval=FALSE}
as.character(3.14) + 1
```

`Error in as.character(3.14) + 1 : non-numeric argument to binary operator`

### Allowed
```{r}
TRUE + 3 # TRUE evaluates to 1; FALSE to 0
```

## Data types

<div class="columns-2">
### Vector
+ A sequence of elements of the **same** basic data type.
```{r}
c(1,FALSE,0.5)
c(1,FALSE,'a')[2]
```
</br>
</br>

### Matrix
+ Two dimensional collection of elements of the **same** basic data type.
```{r}
matrix(
  c(1:9),
  nrow=3, ncol=3,
  byrow=TRUE
)
```
</div>

## Data types
### List
+ A generic vector containing other objects.

```{r}
list.w.vector <- list(
  1.5,TRUE,'a',a=c(1:5))
list.w.vector[[3]]
list.w.vector$a[1]
```

## Data types
### Data frame
+ A list of vectors of equal length used to store data tables.
```{r}
df <- data.frame(
  ALPHABET=letters[1:4],
  INTEGER=c(1:4),
  SQRT=sqrt(c(1:4)),
  GUESSME=c(TRUE,1.2,'0',NA)
)
```


## Data types 
### Data frame
+ A list of vectors of equal length used to store data tables.
```{r eval=FALSE}
df <- data.frame(
  ALPHABET=letters[1:4],
  INTEGER=c(1:4),
  SQRT=sqrt(c(1:4)),
  GUESSME=c(TRUE,1.2,'0',NA)
)
```
```{r echo=FALSE}
df %>%
  kable(align = 'lcccr', digits=3) %>%
  kable_styling(
    full_width = TRUE
  )
```

## Getting help

+ Type `?` followed by a function name.
```{r}
?help
```
+ Or use the search box in `RStudio` under `Help` tab.
+ Or ask the Internet.

<div class="shout-out">
Get used to reading the documentation.
</div>

## Other topics
+ Scalar vs. vectorised functions, i.e., functions that work on a single value vs. al values in a vector.
+ Family of `apply` functions.
+ User-defined functions.
```{r}
sum(1:4)
sapply(df$INTEGER, function(x) { if(x>2) x else -1 })
```

# Data Inspection and Manipulation

## Import data

### `read.csv`
+ Creates a data frame from a csv file.
```{r eval=FALSE}
df <- read.csv('./path/to/file.csv', sep=",", header=TRUE, ...)
```

### `readLines`
+ Reads text lines from a connection.
+ Doesn't have to be a csv.
+ Useful when you don't want to read only a part of a large text file to prevent memory outage.
```{r eval=FALSE}
readLines('./path/to/file.txt', n=-1L) # read all lines
readLines('./path/to/file.txt', n=5) # read first 5 lines
```

## Import data
### Example
+ Data retreived from [BuzzFeed Github Repository (https://github.com/BuzzFeedNews/2015-12-mass-shooting-intervals)](https://github.com/BuzzFeedNews/2015-12-mass-shooting-intervals).
```{r}
df <- read.csv('data/shooting-usa.csv')
```
```{r echo=FALSE}
head(df) %>%
  kable(align = 'rccl') %>%
  kable_styling(
    full_width = TRUE
  )
```

## Basic inspection
### `head` and `tail`
+ Displays first or last parts of data.
```{r}
head(df)
```

## Basic inspection
### `dim`
+ Returns the number of rows x the number of columns.
+ Use `length` for vectors and lists.
```{r}
dim(df)
```

## Basic inspection
### `str`
+ Displays the internal structure of the object.
```{r}
str(df)
```

## Basic inspection
### `summary`
+ Displays basic summaries for each column.
```{r}
summary(df)
```

## Basic manipulation
### Data class conversion
+ `Date` is coded as a `Factor`.
+ `Factor` variables are categorical.
+ We want `Date` to represent...dates.
```{r eval=FALSE}
as.Date(vector, 'Format')
```

## Basic manipulation
### Data class conversion
```{r}
df$Location <- as.character(df$Location)
df$Date <- as.Date(df$Date, '%m/%d/%Y')
str(df)
```

## Basic manipulation
### Data class conversion
+ Allows methods specific to `Date` class.
```{r}
df$Year <- as.numeric(format(df$Date,'%Y'))
df$Month <- months(df$Date)
df$Week.Day <- weekdays(df$Date)
df$Date.Ind <- julian(df$Date, as.Date('2013-01-01'))
head(df[c('Date', 'Year', 'Month', 'Week.Day', 'Date.Ind')])
```

## Basic manipulation
### Data class conversion
+ `Month` and `Week.Day` can be coded as `Factor`s. 
```{r}
df$Month <- factor(
  df$Month, 
  c('January', 'February', 'March', 'April', 'May', 'June',
    'July', 'August', 'September', 'October', 'November', 'December'))
df$Week.Day <- factor(
  df$Week.Day, c('Sunday', 'Monday', 'Tuesday', 
                 'Wednesday', 'Thursday', 'Friday', 'Saturday'))
```

## Basic manipulation
### Text manipulation
+ Part of text can be extracted by pattern.
```{r}
head(df$Location)
```

## Basic manipulation
### Text manipulation
+ Part of text can be extracted by pattern.
```{r}
df$State <- sub('.*, ?', '', df$Location)
unique(df$State)
```

## Basic manipulation
### Text manipulation
+ And more...
```{r}
df$State <- toupper(df$State)
df$State[grep('D.?C.?', df$State)] <- 'DC'
df$State[grep('ILLINOIS', df$State)] <- 'IL'
df$State[grep('KANSAS', df$State)] <- 'KS'
df$State[grep('LOUISIANA', df$State)] <- 'LA'
df$State[grep('PUERTO', df$State)] <- 'PR'
df$State[grep('TENNESSEE', df$State)] <- 'TN'
sort(unique(df$State))
```

## Basic manipulation
### `rbind` and `cbind`
+ Combine R Objects (e.g., data frames) by rows or columns.
```{r eval=FALSE}
# Example
df.2013 <- read.csv('path/to/file2013.csv')
df.2014 <- read.csv('path/to/file2014.csv')
df.2015 <- read.csv('path/to/file2015.csv')
# combine
df.all <- rbind(
  df.2013, df.2014, df.2015
)
```

# Basic Plotting
## Histogram
```{r eval=FALSE}
hist(value.vector, ...)
```

## Scatter plot 
```{r eval=FALSE}
plot(x.vector, y.vector, col=color, ...)
points(x.vector, y.vector, col=color, ...)
```
```{r echo=FALSE}
par(bg=NA) # transparent background for presentation.
plot(df$Injured, df$Dead, 
     col=alpha(brewer.pal(8, 'Accent')[2], 0.5),
     pch=16, axes=FALSE, xlim=c(0,20), ylim=c(0,15),
     xlab='Injuries', ylab='Fatalities')
axis(1, c(0, 10, 20))
axis(2, c(0, 5, 10, 15))
mtext('Number of fatalities and injuries per shooting incident in the U.S. between 2013 and 2015.', adj=0)
```

## Line chart
```{r eval=FALSE}
plot(x.vector, y.vector, type='l', col=color, ...)
lines(x.vector, y.vector, col=color, ...)
```
```{r echo=FALSE}
# sum by month
df$Year.Month <- format(df$Date,'%Y-%m')
incident.by.ym <- table(df$Year.Month)
injury.by.ym <- tapply(df$Injured, df$Year.Month, sum)
dead.by.ym <- tapply(df$Dead, df$Year.Month, sum)
# plot
par(bg=NA)
plot(injury.by.ym, col=2, type='l', ylim=c(0,200), axes=FALSE)
lines(c(incident.by.ym), col=8)
lines(dead.by.ym, col=6)
axis(1, c(-5, 1, 13, 25, 37, 50), c('','2013', '2014', '2015', '2016', ''))
axis(2, c(0, 100, 200))
legend(
  0, 200, c('Incident count', 'Injury count', 'Fatality count'),
  bty='n', col=c(8, 2, 6), lty=1, cex=0.7)
mtext('Fatalities and injuries from shooting incidents in the U.S per month over time.',
      adj=0)
```

## Bar plot
```{r eval=FALSE}
barplot(height.vector, ...)
```
```{r echo=FALSE}
par(bg=NA)
xc <- barplot(table(df$Month), border=NA, 
              names.arg=substr(levels(df$Month),0,3),
              cex.names=0.7, ylim=c(0,200), axes=FALSE)
lines(xc, tapply(df$Injured, df$Month, sum)/c(rep(3,11),2), col=2)
points(xc, tapply(df$Injured, df$Month, sum)/c(rep(3,11),2), pch=16, col=2)
lines(xc, tapply(df$Dead, df$Month, sum)/c(rep(3,11),2), col=6)
points(xc, tapply(df$Dead, df$Month, sum)/c(rep(3,11),2), pch=16, col=6)
axis(2, c(0, 100, 200), las=2)
mtext('Fatalities and injuries from shooting incidents in the U.S per month.',
      adj=0)
```

# Model Fitting
## Linear regression
```{r eval=FALSE}
fitted <- lm(formula=y~x1+x2, data=df, ...)
summary(fitted)
```

## Example
### Generalized additive model
$$ \rm{Casualties}_i \sim \mathcal{N}(\mu_i, \sigma^2) \\
\mu_i = \beta_0+\rm{Month}_i\beta + f(\rm{Week}_i,\nu) $$

## Example
```{r message=FALSE}
library(mgcv)
df$Casualty <- df$Injured + df$Dead
df$Week.Ind <- floor(
  julian(df$Date, as.Date('2013-01-01'))/7
)
df.week <- aggregate(df$Casualty, list(Week=df$Week.Ind), sum)
names(df.week) <- c('Week', 'Count')
df.week$Date <- as.Date(as.POSIXct(
  df.week$Week*7*24*60*60, origin = "2013-01-01", tz="UTC"))
df.week$Month <- factor(
  months(df.week$Date), 
  c('January', 'February', 'March', 'April', 'May', 'June',
    'July', 'August', 'September', 'October', 'November', 'December'))
fitted <- gam(Count ~ s(Week) + Month, data=df.week)
```

## Example
```{r echo=FALSE}
par(bg=NA)
week.seq <- data.frame(
  Date=seq(min(df$Date), as.Date('2017-01-01'), by='week'))
week.seq$Week <- floor(julian(week.seq$Date, as.Date('2013-01-01'))/7)
week.seq$Month <- factor(
  months(week.seq$Date), 
  c('January', 'February', 'March', 'April', 'May', 'June',
    'July', 'August', 'September', 'October', 'November', 'December'))
week.pred <- predict.gam(fitted, week.seq, se.fit=TRUE)
week.seq$fit <- week.pred$fit
week.seq$lwr <- week.pred$fit - qnorm(0.975)*week.pred$se.fit
week.seq$upr <- week.pred$fit + qnorm(0.975)*week.pred$se.fit
# plot time trend
plot(week.seq$Week, week.seq$fit, col=1, type='l',
     xlab='', ylab='Weekly shooting casualty counts', 
     axes=FALSE, ylim=c(0,100))
# shade 95% CI
polygon(c(rev(week.seq$Week), week.seq$Week),
        c(rev(week.seq$lwr), week.seq$upr),
        border=NA, col=alpha(8, 0.2))
points(df.week$Week, df.week$Count, pch=16, col=alpha(7, 0.5))
axis(1, julian(
  as.Date(c(
    '2013-01-01', '2014-01-01', '2015-01-01', '2016-01-01',
    '2017-01-01')), 
  as.Date('2013-01-01'))/7, 
  labels=c('2013', '2014', '2015', '2016', '2017'))
axis(2, c(0, 50, 100), las=2)
mtext('Estimated weekly casualty counts from shooting incidents in the U.S.', 
      adj=0)
```

## Example
```{r echo=FALSE}
par(bg=NA)
month.seq <- data.frame(
  Date = seq(as.Date('2013-01-01'), by='months', length.out=12))
month.seq$Week <- floor(julian(
  month.seq$Date, as.Date('2013-01-01'))/7)
month.seq$Month <- factor(
  months(month.seq$Date), 
  c('January', 'February', 'March', 'April', 'May', 'June',
    'July', 'August', 'September', 'October', 'November', 'December'))
month.pred <- predict.gam(fitted, month.seq, se.fit=TRUE, type='terms', terms='Month')
month.seq$fit <- month.pred$fit
month.seq$lwr <- month.pred$fit - qnorm(0.975)*month.pred$se.fit
month.seq$upr <- month.pred$fit + qnorm(0.975)*month.pred$se.fit
# plot seasonality
plot(month.seq$Week, month.seq$fit, col=1, pch=16,
     xlab='', ylab='Change in weekly casualty counts',
     ylim=c(-10, 50), axes=FALSE)
lines(month.seq$Week, month.seq$fit, col=1)
# shade 95% CI
polygon(c(rev(month.seq$Week), month.seq$Week),
        c(rev(month.seq$lwr), month.seq$upr),
        border=NA, col=alpha(8, 0.2))
axis(1, month.seq$Week, labels=substr(month.seq$Month,0,3), cex.axis=0.7)
axis(2, c(-10, 0, 25, 50))
mtext('Estimated monthly change in weekly casualty counts\nfrom shooting incidents in the U.S.', adj=0)
```

## Example
```{r echo=FALSE}
par(bg=NA)
week.seq <- data.frame(
  Date=seq(min(df$Date), max(df$Date), by='week'))
week.seq$Week <- julian(week.seq$Date, as.Date('2013-01-01'))/7
week.seq$Month <- factor(
  months(week.seq$Date), 
  c('January', 'February', 'March', 'April', 'May', 'June',
    'July', 'August', 'September', 'October', 'November', 'December'))
week.pred <- predict.gam(fitted, week.seq, se.fit=TRUE,
                       type='terms', terms='s(Week)')
week.seq$fit <- week.pred$fit + min(week.pred$fit)
week.seq$lwr <- week.pred$fit + min(week.pred$fit) - qnorm(0.975)*week.pred$se.fit
week.seq$upr <- week.pred$fit + min(week.pred$fit) + qnorm(0.975)*week.pred$se.fit
# plot time trend
plot(week.seq$Week, week.seq$fit, col=1, type='l',
     xlab='', ylab='Change in shooting incident counts',
     axes=FALSE, ylim=c(-7, 7))
# shade 95% CI
polygon(c(rev(week.seq$Week), week.seq$Week),
        c(rev(week.seq$lwr), week.seq$upr),
        border=NA, col=alpha(8, 0.2))
axis(1, floor(julian(
  as.Date(c('2013-01-01', '2014-01-01', '2015-01-01', '2016-01-01')), 
  as.Date('2013-01-01'))/7), 
  labels=c('2013', '2014', '2015', '2016'))
axis(2, c(-7, 0, 7), las=2)
mtext('Estimated trend in weekly casualty counts\nfrom shooting incidents in the U.S.', 
      adj=0)
```

# Resources
## Resources
+ [Codes for the workshop](https://github.com/mjmoon/r_workshop)
+ [Search engien for `R` related stuff](https://rseek.org/)
+ [`R` homepage](https://www.r-project.org/)
+ [http://www.r-tutor.com/r-introduction](http://www.r-tutor.com/r-introduction)
+ [Tutorial on R apply family](https://www.datacamp.com/community/tutorials/r-tutorial-apply-family)
